% discussion.tex
We have shown that pattern multiplicity in the modern Hopfield energy provides a principled,
training-free mechanism for conditioning protein sequence generation on functional subsets. The
theoretical framework is exact: multiplicity weights tilt the Boltzmann distribution through a
simple logit bias, and the resulting Langevin dynamics sample from a known distribution with no
approximation. A single scalar $\rho$ continuously interpolates between unconditioned generation
and hard subset curation, and the practitioner can compute the required $\rho$ for any target
effective designated fraction directly from the family composition (Eq.~\ref{eq:rho-inverse}). On
the Kunitz domain, curated memory achieved 100\% phenotype transfer with as few as three
designated input sequences; multiplicity-weighted generation on the full family increased
phenotype fidelity from the natural proportion (32\%) to 63\% at $\rho = 500$; and the
cross-family validation on five Pfam families showed a quantitative relationship between
PCA-space geometry and conditioning effectiveness.

The central finding is the calibration gap and its decomposition. The gap is not a
failure of the multiplicity-weighted energy (the attention weights track $f_{\mathrm{eff}}$ with
$< 0.3\%$ deviation across all families and $\rho$ values tested) but a well-characterized
property of the PCA encoding that compresses $20L$-dimensional sequence space into $d$ principal
components. When the designated and background subsets overlap in PCA space (low $S$), the
energy-level tilt does not propagate to the decoded residue identity; when they separate
(high $S$), it does. The cross-family validation across five Pfam families produced a monotonic
negative relationship between $S$ and $\Delta$ ($R^{2} = 0.80$ for a linear fit), indicating
that the separation index is a useful \emph{a priori} predictor of conditioning effectiveness.
This yields a practitioner's decision rule: compute $S$ (a cheap operation requiring only PCA and
pairwise cosine similarities, no generation); if $S > 0.3$, multiplicity weighting preserves
fold-level constraints from the full family while achieving high phenotype transfer (SH3 and
Homeobox domains, $\Delta < 0.05$); if $S < 0.2$, hard curation is necessary because the PCA
bottleneck attenuates the signal (WW and Forkhead domains, $\Delta = 0.27$--$0.64$); the
intermediate regime ($0.2 < S < 0.3$, Kunitz domains) admits a tradeoff controlled by
$(\rho,\, \beta/\beta^{*})$.

Hard curation achieves complete phenotype transfer with as few as three designated
input sequences. Many experimental campaigns (phage display, yeast
surface display, functional selections, or even literature curation) produce exactly this kind of
small, functionally characterized set. The practitioner brings external knowledge (``these
sequences bind my target,'' ``these variants are thermostable,'' ``these hits came from my
screen'') and SA converts this small set into a large, diverse candidate library while preserving
the input phenotype. This ``small-set amplifier'' capability complements high-throughput screening:
the screen identifies a few hits, SA expands them computationally, and the expanded library can be
fed back into a second round of selection or directly into functional assays. The method itself is
agnostic to the meaning of the designation; it biases the sampler toward the designated patterns
regardless of whether the underlying phenotype is binding, stability, expression, or any other
property. The scaling study shows that even $K_{\mathrm{des}} = 3$ suffices for phenotype
transfer; diversity improves with more input sequences but saturates near
$K_{\mathrm{des}} = 20$.

The $\omega$-conotoxin experiment extends the small-set amplifier result to a therapeutically
relevant peptide family and illustrates a practically important design choice: whether to seed SA
from the full family or from a curated functional subset. Full-family seeding is appropriate when
the goal is to generate novel family members with broad structural diversity; curated seeding is
necessary when the goal is to preserve a specific pharmacophore. In the $\omega$-conotoxin case,
the O-superfamily is structurally coherent (all members share the C--C--CC--C--C disulfide
framework) but functionally heterogeneous (members block N-type, P/Q-type, or sodium channels
with varying selectivity). Full-family seeding dilutes the Cav2.2-specific Tyr pharmacophore
because the PCA encoding reflects the full structural diversity; curated seeding from the 23
N-type blockers concentrates the PCA subspace on Cav2.2-relevant variation, allowing SA to
regenerate it with 98\% fidelity. This generalizes the separation-index interpretation: even
when $S$ is moderate, hard curation can achieve high pharmacophore transfer by
restricting the memory to the functional subset of interest. The cost (lower sequence diversity
and reduced novelty, 0.46 vs.\ 0.60) reflects the narrower PCA basis, consistent with the
Kunitz scaling study.

The AlphaFold2-multimer analysis shows that SA-generated $\omega$-conotoxins are
indistinguishable from natural sequences under complex prediction: all four metrics (pTM, iPTM,
confidence, interface pLDDT) overlap between generated and control groups with no significant
differences. This is consistent with the structural and pharmacophore validation: generation
preserves the properties that determine predicted binding geometry. The absolute scores are
modest across all groups, a known limitation of multimer prediction for small-peptide--channel
complexes; experimental binding assays are needed to determine whether the preserved
pharmacophore translates to functional Cav2.2 block.

Several limitations should be acknowledged. The cross-family regression spans five Pfam families
(plus $\omega$-conotoxin) with $R^{2} = 0.80$ for a linear fit; while the monotonic trend is
clear, additional families would help characterize the functional form of the $S$--$\Delta$
relationship, which may be nonlinear (the gap appears to decay steeply below $S \approx 0.3$
and flatten near zero above it). Our functional splits are defined by single marker positions
(P1 in Kunitz, Trp in SH3, specificity-loop residue in WW, Gln at position~50 in Homeobox,
H/N at the recognition helix in Forkhead); multi-position phenotypes, where specificity depends
on coordinated residue identities at several positions, may exhibit different calibration
behavior. We have not experimentally
validated the binding activity of generated sequences; the phenotype fidelity reported here is a
necessary but not sufficient condition for functional binding. For the $\omega$-conotoxin family,
we validated SA-generated sequences against published SAR data from alanine-scanning and
site-directed mutagenesis studies~\cite{kimTyr13Essential1995,satoBasicResidues1993,lewisConotoxinSAR2012}
and found that strong-seeded generation preserved all experimentally identified binding
determinants (Table~\ref{tab:sar-agreement}); for the Kunitz domain, deep mutational scanning
data for the BPTI--trypsin interaction are now
available~\cite{heyneBPTIDMS2021} and could provide a quantitative binding-landscape comparison
in future work. Computational binding affinity predictions (e.g., via AlphaFold-Multimer or
FoldX) could provide an additional intermediate validation tier between sequence-level markers
and experimental assays. Finally, we tested interface-weighted PCA
encoding as a strategy to close the calibration gap, but found it neutral at low weights and
destructive at high weights; the PCA bottleneck reflects the geometric entanglement of functional
and structural variation in the family, not the absence of interface-position components.

Looking forward, the multiplicity framework opens several directions. The binary split
(designated/background) generalizes naturally to multi-class conditioning by assigning different
multiplicities to different functional categories. The PCA bottleneck that limits conditioning
effectiveness might be reduced by alternative sequence encodings. Learned representations from
protein language models, for instance, may better separate functional subsets than linear PCA.
The $\beta$ lever suggests that adaptive temperature schedules, increasing $\beta$ during sampling
to sharpen the multiplicity bias, could improve phenotype transfer without permanently sacrificing
diversity. These extensions remain within the analytic, training-free framework that makes
stochastic attention practical for the scarce-data regime.
