% introduction.tex
Generating novel protein sequences that fold correctly and perform a desired function is a
central goal of computational protein engineering. Several classes of methods now address this
problem. Deep generative models, including variational autoencoders~\cite{sinaiAdaptiveMLDE2020},
autoregressive language models~\cite{madaniProGen2023}, and diffusion
models~\cite{alamdariProteinGeneration2023}, learn sequence distributions from large databases
and can produce diverse candidates, but require substantial training data and GPU resources.
Structure-conditioned inverse folding methods~\cite{dauparasProteinMPNN2022} design sequences for
a given backbone but require a target structure as input. Protein language
models~\cite{esmfold2022} trained on evolutionary databases can score or sample sequences, but
conditioning them on a specific functional property typically requires fine-tuning on labeled
data. Profile HMMs~\cite{hmmer3} capture family-level statistics from alignments but, as we show
below, produce sequences with low structural confidence and poor compositional fidelity. For the
scarce-data regime (families with tens to hundreds of members, typical of Pfam seed alignments),
most learned models lack sufficient training signal.

Stochastic attention (SA) was developed for this regime~\cite{varnerSAProtein2026}. SA treats the
energy function of a modern Hopfield network, constructed directly from a small family alignment,
as a Boltzmann density and samples from it using Langevin dynamics. The method requires no
training, runs on a laptop in seconds, and produces sequences that preserve family-level amino
acid composition, per-position conservation, and predicted three-dimensional fold. A key
limitation, however, is that SA treats all sequences in the alignment equally. Every family
member contributes the same weight to the energy landscape, so the sampler explores the full
family distribution without preference for any functional subset. In practice, experimentalists
often want something more specific: not just ``a plausible Kunitz domain,'' but a Kunitz domain
that \emph{inhibits trypsin}; not just ``a plausible SH3 domain,'' but one that binds a
particular proline-rich motif. Family alignments define a shared fold (all 99 Kunitz sequences
share the same disulfide framework and double-loop architecture), but the members differ in
which target they bind, and that specificity is encoded in a small number of positions (e.g.,
the P1 contact residue) rather than in the global fold. The alignment captures structure; the
practitioner wants to condition on function. The question is: \emph{how can we condition the
generative process on a functional subset without retraining?}

A seemingly simple answer is to \emph{curate} the input, restricting the alignment to only those
sequences that a practitioner has identified (through experimental screens, literature annotation,
or functional assays) as belonging to the functional subset of interest. SA then generates new
sequences in their image. We show that this strategy works: even three designated input sequences
suffice for complete phenotype transfer in the Kunitz domain. But hard curation discards all
remaining family members, losing the fold-level constraints they provide. A more principled
approach would retain the full family while continuously upweighting the designated subset.

SA works by encoding each sequence in the alignment as a column in a matrix (the memory matrix),
then running a physics-inspired sampler that generates new sequences by taking noisy steps through
the landscape defined by these stored sequences. At each step, the sampler computes an attention
distribution over the stored sequences and moves toward a weighted average of them. The key
observation we exploit is that adding a bias term to these attention scores deepens the energy
wells around specific sequences, causing the sampler to spend more time near them. Concretely,
we assign a multiplicity weight $r_k > 0$ to each stored sequence and add $\log r_k$ to its
attention score before the softmax normalization. A single parameter, the multiplicity ratio
$\rho = r_{\mathrm{designated}}/r_{\mathrm{background}}$, controls the strength of this bias:
$\rho = 1$ recovers standard SA (no preference), while increasing $\rho$ shifts generation
toward the designated subset. No sequences are duplicated, no parameters are trained, and no
architectural changes are needed. The method is agnostic to what the designated subset
represents (binding, thermostability, organism of origin); it simply biases the sampler toward
the sequences the user points to.

In this study, we developed the theory of multiplicity-weighted SA and tested it on five Pfam
protein families (Kunitz, SH3, WW, Homeobox, and Forkhead domains) spanning a range of family
sizes ($K = 55$--$420$) and functional-subset geometries. We showed that the multiplicity
weights enter the sampler as a simple logit bias, preserving the analytic score function and
$\mathcal{O}(dK)$ cost of the original method. The energy-level conditioning is exact: the
softmax attention on designated patterns matches the target fraction to within $0.3\%$ at all
multiplicity ratios tested. However, we found that the decoded sequence phenotype can fall short
of this target, because the PCA encoding that compresses sequences into the sampler's state space
does not always preserve the residue-level variation that defines the functional split. We termed
this discrepancy the calibration gap and showed that it is predicted by a simple geometric
quantity, the Fisher separation index $S$, which measures how well PCA separates designated from
background sequences. Across five families spanning a fourfold range of $S$, the gap decreased
monotonically with increasing separation ($R^{2} = 0.80$). This provides a practitioner's
decision rule: compute $S$ before generating any sequences; if $S$ is high, multiplicity
weighting transfers the phenotype directly; if $S$ is low, use hard curation instead.
