% theory.tex -- Theory subsections under Materials and Methods
\subsection{From Hopfield Energy to Multiplicity-Weighted Score Function}\label{sec:theory}

Let $\vX = [\vm_1, \ldots, \vm_K] \in \R^{d \times K}$ be a memory matrix whose columns are
unit-norm, PCA-encoded protein sequences from a family alignment of $K$ members, each represented
in $d$ dimensions. The modern Hopfield energy~\cite{ramsauerHopfieldNetworksAll2021} at state
$\vxi \in \R^d$ is $E(\vxi) = \tfrac{1}{2}\|\vxi\|^{2} - \tfrac{1}{\beta}\log\sum_{k}\exp(\beta\,\vm_{k}^{\!\top}\vxi)$, where $\beta > 0$ is an inverse temperature. The score function
$\nabla_{\vxi}\log p_{\beta}(\vxi) = \vX\,\softmax(\beta\,\vX^{\!\top}\vxi) - \vxi$ is
\emph{exact} and computed by a single attention operation; applying the unadjusted Langevin
algorithm (ULA) with step size $\alpha \in (0,1)$ yields the stochastic attention
update~\cite{varnerSAProtein2026}:
\begin{equation}\label{eq:ula}
  \vxi_{t+1}
    = (1-\alpha)\,\vxi_{t}
    + \alpha\,\vX\,\softmax\!\bigl(\beta\,\vX^{\!\top}\vxi_{t}\bigr)
    + \sqrt{2\alpha/\beta}\;\veps_{t},
  \qquad \veps_{t} \sim \mathcal{N}(\mathbf{0},\,\mathbf{I}_{d}).
\end{equation}
We extend this framework to condition generation on a functional subset by assigning a
multiplicity weight $r_k > 0$ to each stored pattern. The \emph{multiplicity-weighted} Hopfield
energy is
\begin{equation}\label{eq:weighted-energy}
  \boxed{
  E_{\vr}(\vxi) = \tfrac{1}{2}\|\vxi\|^{2}
    - \tfrac{1}{\beta}\log\sum_{k=1}^{K} r_k\,\exp\!\bigl(\beta\,\vm_{k}^{\!\top}\vxi\bigr),}
\end{equation}
where $\vr = (r_1, \ldots, r_K)^{\!\top}$.
When $r_k = 1$ for all $k$, Eq.~\eqref{eq:weighted-energy} reduces to the standard energy.
Patterns with higher multiplicity deepen the energy wells in their vicinity; this is equivalent
to storing $r_k$ copies of pattern $\vm_k$ without the $\mathcal{O}(d \cdot \sum_k r_k)$
memory cost.

\begin{proposition}[Score function]\label{prop:score}
The score function of $p_{\vr}(\vxi) \propto \exp(-\beta\,E_{\vr}(\vxi))$ is
$\nabla_{\vxi}\log p_{\vr}(\vxi) = \vX\,\softmax(\beta\,\vX^{\!\top}\vxi + \log\vr) - \vxi$,
where $\log\vr = (\log r_1, \ldots, \log r_K)^{\!\top}$ is the elementwise log-multiplicity.
\end{proposition}

The proof follows from the identity $r_k\,\exp(a_k) = \exp(a_k + \log r_k)$: the weights are
absorbed into the softmax logits. The resulting Langevin update is
\begin{equation}\label{eq:weighted-ula}
  \boxed{
  \vxi_{t+1}
    = (1-\alpha)\,\vxi_{t}
    + \alpha\,\vX\,\softmax\!\bigl(\beta\,\vX^{\!\top}\vxi_{t} + \log\vr\bigr)
    + \sqrt{2\alpha/\beta}\;\veps_{t}.}
\end{equation}
The \emph{only} change from Eq.~\eqref{eq:ula} is the addition of $\log r_k$ to each logit
before the softmax. The memory matrix $\vX$ is unchanged, the noise schedule is unchanged, and
the sampler retains its $\mathcal{O}(dK)$ per-step cost.

\subsection{Multiplicity Ratio and Phase Transition}\label{sec:phase}

For a binary functional split (designated subset vs.\ background), we parameterize the multiplicity
vector by a single scalar $\rho = r_{\mathrm{designated}}/r_{\mathrm{background}}$, setting
$r_k = \rho$ for patterns in the user-designated subset and $r_k = 1$ otherwise. The effective
designated fraction (the share of the softmax probability mass attributable to designated
patterns when all patterns have equal similarity to $\vxi$) is
$f_{\mathrm{eff}}(\rho) = K_{\mathrm{des}}\rho/(K_{\mathrm{des}}\rho + K_{\mathrm{bg}})$, where
$K_{\mathrm{des}}$ and $K_{\mathrm{bg}} = K - K_{\mathrm{des}}$ are the number of designated and
background patterns. At $\rho=1$,
$f_{\mathrm{eff}} = K_{\mathrm{des}}/K$; as $\rho \to \infty$, $f_{\mathrm{eff}} \to 1$. Inverting
gives the multiplicity ratio needed for a target fraction $f$:
\begin{equation}\label{eq:rho-inverse}
  \rho(f) = \frac{f \cdot K_{\mathrm{bg}}}{K_{\mathrm{des}}\,(1-f)}.
\end{equation}
The standard SA framework exhibits a phase transition at critical inverse temperature $\beta^{*}$
that depends on the PCA dimension $d$ and the number of stored patterns. Under multiplicity
weighting, the relevant quantity is the effective number of patterns
$K_{\mathrm{eff}}(\vr) = (\sum_k r_k)^{2} / \sum_k r_k^{2}$, which equals $K$ when all $r_k$
are equal and converges to $K_{\mathrm{des}}$ as $\rho \to \infty$.

\begin{remark}[Phase transition: analytic baseline and multiplicity shift]\label{rem:beta-star}
The companion paper~\cite{varnerSAProtein2026} establishes, via a concentration-of-measure
argument on the softmax attention entropy, that for the \emph{unweighted} SA sampler ($\rho=1$)
the phase transition satisfies
\begin{equation}\label{eq:beta-star}
  \beta^{*} \;\approx\; 1.57 + 0.28\sqrt{d},
\end{equation}
with $R^{2}=0.97$ across eight Pfam families spanning $d \in [18, 186]$, and is nearly
independent of $K$ for $K \geq 30$.
At $\rho=1$, Eq.~\eqref{eq:beta-star} therefore gives the analytic value directly from the
alignment. For $\rho > 1$, the logit biases $\log r_k$ pre-concentrate the attention
distribution at $\beta=0$: the weighted entropy at zero temperature is $H_{\vr}(0) = \log
K_{\mathrm{eff}}(\vr) < \log K$, a lower starting entropy than the unweighted case.
Reaching the retrieval-phase inflection from this pre-concentrated starting point requires
a higher inverse temperature, so $\beta^{*}(\rho) \geq \beta^{*}(1)$.
For intermediate $\rho$, $\beta^{*}(\rho)$ is located as the inflection of the weighted
attention entropy $H_{\vr}(\beta)$ via a one-dimensional sweep, a negligible computation
relative to sequence generation.
\end{remark}

Empirically, $\beta^{*}$ increases monotonically from 4.4 ($\rho = 1$, $K_{\mathrm{eff}} = 99$)
to 9.3 ($\rho = 1000$, $K_{\mathrm{eff}} = 32.1$) in the Kunitz domain, consistent with
Remark~\ref{rem:beta-star}: more inverse temperature is required as the logit biases concentrate
probability mass on fewer effective patterns.

\subsection{The Calibration Gap}\label{sec:gap}

The signal chain from multiplicity weights to decoded sequence phenotype passes through several
transformations:
\begin{equation}\label{eq:signal-chain}
  \rho
  \;\xrightarrow{\;\text{exact}\;}\;
  f_{\mathrm{eff}}
  \;\xrightarrow{\;\text{exact}\;}\;
  \bar{a}_{\mathrm{des}}
  \;\xrightarrow{\;\text{lossy}\;}\;
  f_{\mathrm{soft}}
  \;\xrightarrow{\;\text{nonlinear}\;}\;
  f_{\mathrm{obs}},
\end{equation}
where $\bar{a}_{\mathrm{des}}$ is the mean attention weight on designated patterns during
sampling, $f_{\mathrm{soft}}$ is the continuous-valued residue probability at the marker position,
and $f_{\mathrm{obs}}$ is the fraction of generated sequences displaying the designated phenotype
after argmax decoding. We define the calibration gap $\Delta = f_{\mathrm{eff}} - f_{\mathrm{obs}}$
and decompose it into three independently measurable components:
\begin{equation}\label{eq:gap-decomp}
  \Delta = \underbrace{(f_{\mathrm{eff}} - \bar{a}_{\mathrm{des}})}_{\Delta_{\mathrm{attn}}\;\approx\;0}
  + \underbrace{(\bar{a}_{\mathrm{des}} - f_{\mathrm{soft}})}_{\Delta_{\mathrm{PCA}}}
  + \underbrace{(f_{\mathrm{soft}} - f_{\mathrm{obs}})}_{\Delta_{\mathrm{argmax}}}.
\end{equation}
To predict the PCA gap across families, we define the Fisher separation index
$S = (\bar{c}_{\mathrm{within}} - \bar{c}_{\mathrm{between}}) /
[\tfrac{1}{2}(\sigma_{\mathrm{within}} + \sigma_{\mathrm{between}})]$, where $\bar{c}$ and
$\sigma$ denote the mean and standard deviation of pairwise cosine similarities within and
between functional groups in PCA space.
