\documentclass{article}
\PassOptionsToPackage{numbers, compress}{natbib}
\usepackage[preprint]{neurips_2026}

% --- standard packages ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[colorlinks=true,linkcolor=blue!70!black,citecolor=blue!70!black,urlcolor=blue!70!black]{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{bm}
\usepackage{multirow}

% --- figure/table placement ---
\setcounter{topnumber}{1}
\setcounter{bottomnumber}{1}
\renewcommand{\topfraction}{0.5}

% --- theorem environments ---
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% --- custom commands ---
\newcommand{\vxi}{\bm{\xi}}
\newcommand{\vX}{\mathbf{X}}
\newcommand{\vm}{\mathbf{m}}
\newcommand{\vr}{\mathbf{r}}
\newcommand{\veps}{\bm{\epsilon}}
\newcommand{\R}{\mathbb{R}}
\DeclareMathOperator{\softmax}{softmax}

\title{Conditioning Protein Sequence Generation via Pattern Multiplicity\\in the Modern Hopfield Energy}

\author{%
  Jeffrey D.~Varner \\
  R.F. Smith School of Chemical and Biomolecular Engineering\\
  Cornell University, Ithaca, NY 14850 \\
  \texttt{jdv27@cornell.edu} \\
}

\begin{document}

\maketitle

\begin{abstract}
\input{sections/abstract}
\end{abstract}

\noindent\fbox{\parbox{\textwidth}{\small\textbf{Significance Statement.}
Protein engineers routinely isolate a handful of functional binders---via phage display, yeast
surface display, or functional selections---but lack a systematic way to expand these small sets
into diverse candidate libraries without retraining a generative model. We show that a single
scalar parameter added to the softmax logits of a Hopfield-energy sampler continuously conditions
generation from ``any family member'' to ``only binders,'' with no training, no GPU, and no
architectural changes. The effectiveness of this conditioning is predicted \emph{a priori} by a
cheap geometric property of the alignment, giving practitioners a decision rule before any
sequences are generated.}}

\bigskip

\section{Introduction}
\input{sections/introduction}

\section{Results}
\input{sections/results}

\section{Discussion}
\input{sections/discussion}

\section{Materials and Methods}
\input{sections/theory}
\input{sections/methods}

\section*{Acknowledgments}
% TODO: Acknowledgments

\paragraph{Author contributions.}
J.D.V.\ conceived the study, developed the theory, wrote the code, performed the experiments, and wrote the paper.

\paragraph{Competing interests.}
The author declares no competing interests.

\paragraph{Data availability.}
All Pfam seed alignments are publicly available from InterPro. Source code and experiment scripts
are available at \url{https://github.com/varnerlab/SA-Binding-Generation-Study}.

\bibliographystyle{unsrtnat}
\bibliography{References_v1}

% --- Figures and Tables (placed after bibliography, PNAS-style) ---
\clearpage

% TODO: Add figures here with [p] full-page placement
% Table 1: Cross-family comparison
% Fig 1: P1 phenotype inheritance (Kunitz)
% Fig 2: Binder scaling (fidelity/diversity/KL)
% Fig 3: rho sweep (4-panel)
% Fig 4: Phase transition entropy curves
% Fig 5: Calibration gap decomposition
% Fig 6: Cross-family calibration + separation vs gap
% Fig 7: Per-family rho sweep panels

% --- SI Appendix ---
\clearpage
\setcounter{figure}{0}
\setcounter{table}{0}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{section}{0}
\renewcommand{\thesection}{S\arabic{section}}

\section*{SI Appendix}
\input{sections/appendix}

\end{document}
